The Causal Effect of News Framing on Opinion Formation

### 2022-06-28 paper/main.tex
Update abstract
Revert "Delete paper directory"
This reverts commit b3ea5c81fc5d1df3344229a23a919c1f6b27542d.

\title{The Effect of News Framing on Issue Attitudes and Voting Behaviour}
\author{Nicolai Berk\footnote{Doctoral Candidate, Dynamics Doctoral Program, Humboldt-Universität zu Berlin, \href{mailto:nicolai.berk@hu-berlin.de}{nicolai.berk@hu-berlin.de}}}
\date{October 2021}

@@

\begin{abstract}
    A large body of literature debates the extent to which media consumption affects consumers' issue attitudes and voting behaviour in contemporary media environments. Studies supporting the minimal effects hypothesis have often used designs estimating the effects of novel, previously unconsumed content on participants' attitudes. However, experimental evidence suggests that media consumers discount the bias of different outlets. Additionally, most studies assume rather than measure the extent of newspaper bias. Combining panel data with state-of-the-art methods of text analysis and exploiting an editorial change, I address both shortcomings with observational data. Results across different specifications indicate that even when assessing within-newspaper changes in issue framing, no effect on attitudes or voting behaviour can be found. These findings add further evidence that media effects in the post-broadcast era are limited.
\end{abstract}

@@

How does news media content affect citizens' issue attitudes? Research on political behaviour has discussed this question for nearly a century. Classic theories of agenda setting hold that the attention devoted to issues in the news media will guide citizens to consider these issues important (\cite{McCombs1972}). Similarly, the literature on framing expects that the presentation of issues in the news media will affect recipients' opinions about these issues (\cite{Nelson1997}). Both theories suggest that media coverage strongly affects citizens' understanding of and opinions about politics.

Yet, a critique of these theories maintains that strong news effects are unlikely in the modern media environment, where consumers self-select into news sources of their liking. As a result, "most media users will rarely find themselves in the path of attitude-discrepant information". Even if consumers had to face such discrepant information, this should reinforce their existing views rather than persuade as they are used to a partisan media diet (\cite[724f]{Bennett2008}). Similarly, agenda setting effects are unlikely as a result of selective exposure (\cite{Lau2021}).

This discussion begs the question: when can we expect media framing effects? Some authors have argued that a lack of media effects can be explained by readers who discount a publications' bias. For example, when a left-wing newspaper endorses a left-wing candidate, this should have no effect on citizens' assessment of the candidate, as the endorsement was to be expected (\cite{Chiang2011a}, see also \cite{Spirig2020}). Additionally, citizens might take cues from their preferred publication because they developed trust in its coverage over time.

@@

\subsection{Case description}

To test this hypothesis, I analyse media and opinion data on the migration issue from the German national election in 2017. Following a period of increased refugee arrivals in Europe in the summer of 2015, the election was dominated by the migration topic and the related upcoming of the radical-right populist challenger party \textit{AfD}. The major two parties dominating German post-war politics, \textit{SPD} and \textit{CDU/CSU}, had to face severe losses following their 'grand coalition' (\cite{Bieber2021, Wessels2021}).

This strong (although still variant, see figure \ref{fig:salience}) salience of the migration issue both in the media as well as in citizens subjective evaluations presents an ideal case to study the effects of shifting media framing on issue attitudes, as continuous coverage on and public debate of the issue are a given.

@@

To measure the dependent variable (issue attitudes) and differentiate the treatment groups (newspaper readers), I employ panel data from the German Longitudinal Election Study (\cite{GLES2019LongTermTracking}). The panel consists of 15 waves, of which 6 (4) contain variables on newspaper consumption and immigration (integration) attitudes. In each wave 9,000-15,000 respondents were interviewed, which allows the precise estimation of effects of media consumption.

Figure \ref{fig:issues} shows the two main dependent variables for each group of readers. Note that higher values indicate more restrictive attitudes towards immigration/integration (see plot subtitles for more information). As can be seen, readers of the tabloid 'Bild' differ strongly from consumers of other outlets, with far more restrictive attitudes. Both variables seem correlated (as expected), with values moving towards the more liberal pole until autumn, where the trend is reversed.

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/salience_papers_focus.png}
    \caption{Share of migration articles out of all articles across time for different newspapers.}
    \label{fig:salience}
\end{figure}

To measure media attention to immigration, I collected 2.5 million newspaper articles from the most important German broadsheets \textit{Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt} and the major German tabloid \textit{Bild} for the period 2013-2019. To identify whether an article is about migration, I train a BERT deep-learning classifier.

As migration content is rare, and my training data needs to be somewhat balanced, I first construct a dictionary of terms related to migration. I use dictionary extension based on German GloVe word-embeddings\footnote{Downloadable from \url{https://deepset.ai/german-word-embeddings}.} to construct a comprehensive dictionary, and apply it to the articles. Based on the relative share of migration words in an article, I draw a stratified sample of 300 articles (100 from the third with fewest migration terms, 100 from the third with most migration terms and 100 from the third in between) for each newspaper for a total of 1,800 papers. Then, a student assistant\footnote{Thank you, Robin. I know it was painful.} hand-coded these articles, assessing whether their main topic is related to migration.

Using this sample, a BERT deep-learning classifier\footnote{\url{https://huggingface.co/bert-base-german-cased}} is fine-tuned on a subset of 1,400 annotated articles. After fine-tuning, the model correctly classifies 95.5\% of the test set (F1: 0.94, recall: 0.93, precision: 0.95). This classifier is then used to annotate all 2.5 million newspaper articles. For the time frame of relevance for this study (the entire year of 2017), out of the total of 388,617 articles, 13,589 (3.5\%) are identified to treat the issue of migration\footnote{Note that the classifier, as well as the following topic model, was trained on a corpus of documents spanning 2013-2019. This was largely a decision to maximise possible applications for the resulting dataset, but also enables to compare issue attention and frames across time for validity checks.}.

Figure \ref{fig:salience} shows the share of migration articles in each newspaper. While the first few months of 2017 show rather high salience of the issue, attention decreases until May, before the share of articles covering this issue increases again in all newspapers except \textit{Bild}, which shows a stark decrease in July and maintains this level afterward. Most newspapers show similar levels of attention to the issue between 2.5\% and 6\% of all articles, with the exception of \textit{SZ}, which moves between 5\% and 12.5\%.

@@

I expect the capital crime, illegal entry/petty crime, and the refugee numbers frames to have negative effects on attitudes towards migrants (as connected to negative considerations regarding immigration), while deportations, internment camps, and drownings in the Mediterranean sea should have positive effects, framing immigration as a humanitarian issue. The labour market frame should show mixed results, given that labour market integration might be perceived as threatening or addressing needs in the German economy. Figure \ref{fig:frames} shows the prevalence of all frames across time for each newspaper. It can be seen that substantial variation exists in the relative emphasis of migration frames both within and across newspapers. Interestingly, some issues seem to vary according to newspaper slant. For example, more right-leaning outlets \textit{Bild, FAZ}, and \textit{Welt} seem to largely ignore the humanitarian crisis in the Mediterranean sea, while they emphasise petty crime more. This lends some first credibility to the expected effects for these frames, although the differences are not clearly visible for the other frames.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/frames_papers_focus.png}
    \caption{Prevalence of different migration frames across newspapers. Gray vertical lines indicate survey field dates.}
    \label{fig:frames}
\end{figure}

@@

I estimate the same DiDs to measure shifts in migration framing in a certain newspaper, controlling for shifts in all other newspapers. To do that, I first sum the attention given to a certain frame in a given newspaper in the past day, week, month, or half year, up to the last survey wave. This allows me to assess whether choosing different lags affects my estimates. Then, I estimate the DiD, indicating how much a newspaper changed its attention to a given migration frame, relative to all others. The DiDs estimating the opinion shifts described in \ref{sec:did_readers} are then matched and regressed on the shifts in migration framing. If increasing attention to a certain frame causes a shift in migration attitudes, these two measures should be highly correlated.

Second, I fit two-way fixed-effect models with individual news framing estimates for each survey interview. I measure exposure to each frame by summing up the frame attention of the past week (relative to the interview) in each newspaper the respondent claimed to have read in the past week and dividing it by the total number of newspapers read. This way, I estimate frame exposure for each individual respondent's interview date and news diet, which allows to estimate the correlation of changes in immigration opinion with the migration framing in recently consumed outlets.

While the DiD's are causally better identified, controlling for shifts in other newspapers and reader groups, they suffer from the fact that surveys are conducted for several days, often up to two weeks, which makes the precise matching of newspaper content impossible. The fixed-effect models control for wave effects and (time-constant) individual characteristics, therefor restricting the explainable variance to everything beyond general trends and individual characteristics, but might include more noise. Another difference is that the DiD-models correlate changes, while the fixed-effect models regress opinion levels on frame attention levels.

@@

\subsection{DiD-assumption: Parallel trends}

@@

\subsection{Distribution of DiD-effect significance}

These highly parallel trends might raise another problem, however. Possibly, no significant variation of reader groups' migration opinion from the general trend can be observed. To address this concern, I estimate DiDs for all wave-issue-paper combinations. This means I use attitudinal questions on all political issues included in the GLES Panel (economic, issues, climate, ...), and assess whether news consumption matters for opinion formation more generally. Figure \ref{fig:p_values} shows the distribution of p-values for the DiD-term from 1,752 fixed-effect models with GLES Panel data. The distribution shows a considerable deviation from the expected uniform distribution (red line) were there no explaining variation in the DiD-term beyond the uniform effects of wave and paper readership. This means, more often than random (12\% vs 5\%), the opinions of consumers of certain news deviate from the opinions of other citizens. This is even more severe for the 192 wave-paper combinations concerning the immigration or integration item used in the remainder of this study, where 18\% of all estimated DiDs fall under the 5\%-threshold\footnote{See plot in appendix \ref{app:p-val_imm}}.  This strongly indicates that news consumption matters for opinion formation and that the data shows variation which can be explained, despite highly parallel trends.

\subsection{Correlation of DiD with frame prevalence}

There seems to be interesting variation in the dependent variable. Can these opinion shifts be explained by shifts in news content? Figure \ref{fig:did_corr} shows the core analysis to estimate the effect of migration framing on reader opinion. The displayed estimates are obtained by regressing the DiD-estimates for issue attitudes on the DiD-estimates for media attention, as described in section \ref{sec:models}. If opinion shifts among newspapers' readers - relative to other readers - can be explained by shifts in newspapers' issue framing, these two variables should be highly correlated.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{paper/vis/effectplot_frames_did.png}
    \caption{Regression coefficients from DiD-models with different specifications.}
    \label{fig:did_corr}
\end{figure}

This is however not the case. Each panel in figure \ref{fig:did_corr} shows the effect of one frame on an attitude variable (controlling for all other frames mentioned in section \ref{sec:frame_selection}), across different lags and treatment definitions. Each dot in this graph represents the coefficient of a model explaining standardised\footnote{Meaning that independent and dependent variables were mean-corrected and re-scaled to standard deviations before regression. A coefficient of one means that a one-standard-deviation shift in the independent variable corresponds to a one-standard deviation shift in the dependent variable.} aggregate opinion shifts with standardised shifts in media attention to specific frames.  For example, in the upper left panel 'crime', we can see two different panes ('immigration' and 'integration') for each dependent variable. The vertical rows distinguish four different lags indicating the frame prevalence of the past day, week, month, half year, or the entire period up to the last survey date. Additionally, the color of the points indicates whether the model was estimated only for readers of a single newspaper (lower N, but no treatment diffusion) or all readers of a newspaper (who might read another newspaper, say SZ readers who also read FAZ).

Only very few estimates are significant and no trends emerge across specifications for any variable. All estimates are clearly clustering around zero, indicating no effect. Estimates of all relationships indicate both positive and negative relationships across specifications. While the 'exclusive reader' estimates show more variation, they also indicate far broader confidence intervals. There is no considerable difference across lags, and the variables only differ by the broader confidence intervals of the integration variable, owing to the fewer waves including this question. Generally, this analysis indicates no effect of media framing of migration on migration attitudes.

Again, most estimates are not significant, differ in direction across specifications, and cluster around 0. While some estimates are unidirectional (crime and labour market in the immigration issue), they show tiny effects (a standard deviation increase in attention to the labour market makes issue attitudes between 0.005 and 0.015 standard deviations more conservative) and are usually not significant. Most importantly, both models do not correspond to expectations. This is probably most visible for the capital and sex crime frame ('capcrime'), which should show strongly positive effects (leading to more restrictive attitudes), but also for the illegal immigration and petty crime frame ('crime'), which consistently shows insignificant negative (i.e. liberalising) effects.

This paper assessed the responsiveness of newspaper consumers' migration attitudes to the migration framing of their respective newspaper. Aggregate difference-in-differences models and individual estimates show that consumers' opinions on immigration and integration were not reacting to changes in the newspaper framing of the respective issue. However, several alternative explanations are possible.

Some other studies found that, instead of reacting with opinion shifts, readers might change their political preferences or mobilise to vote (\cite{Stetka2020}). Similarly, several studies indicate that changes in a media outlet's slant will result in a change in the composition of the consumers (ie those with conflicting attitudes stop reading the newspaper) rather than a change in a fixed readership's attitudes (\cite{Arceneaux2013, Durante2012, Spirig2020}). Whether similar effects can be observed when explicitly measuring newspaper framing are likely next steps for analysis.

Another theoretical alternative might be the idea of \textit{issue evolution} (\cite{Carmines1986}). If a process of 'issue-sorting' took place in 2015, when the migration issue became very salient for the first time in a long time, citizens might have chosen a side and stuck to their opinion afterward. Hence, a second further avenue for analysis might look into opinion stability across a longer time horizon or a different case.

On the empirical side, I cannot ensure that respondents actually received treatment when they say they read the newspaper. Some might only read the sports section, some might only read specific sub-issues (e.g. only the crime news about immigration, but not the news on migrants drowning in the Mediterranean). It also might be that survey responses regarding newspaper consumption are generally more statements of which paper respondents like rather than consume. It is therefor in no way clear whether the treatment actually reaches respondents, although similar studies connecting survey- and media data (\cite{Eberl2015}) have found effects of media content on attitudes and even voting behaviour.

More could be done to measure emphasis frames as well. Recent innovations exploiting word embeddings are particularly noteworthy here, either extracting topics from the embedding space (\cite{Angelov2020, Urman2021}) or measuring conceptual associations across time with diachronic word embeddings (\cite{Rodman2019}).

Beyond the improvement of the analysis, I could try to add a different case and/or exploit a natural experiment. The Bild had an editorial shift in 2017 which seems to have coincided with more emphasis of crime topics - in line with anecdotal evidence of the results of this change among journalists. Returning to an earlier framing of the paper, this case could be presented as additional evidence of null effects. It might also be possible to exploit the increasing attention to the climate issue in 2019 and again in 2020 to assess an opinion formation process from the beginning, although this would necessitate to leave the paper alone for a while, until the survey data is available.

Lastly, if there were in fact no effect of news framing on respondents' attitudes towards migration, there is still a major open question: what does explain the variation that can be observed?

@@

\subsection{P-values immigration and integration issue}\label{app:p-val_imm}

>>>>

\title{The Causal Effect of News Framing on Opinion Formation}
\author{Nicolai Berk\footnote{Doctoral Candidate, Dynamics Doctoral Program, Humboldt-Universität zu Berlin, \href{mailto:nicolai.berk@hu-berlin.de}{nicolai.berk@hu-berlin.de}}}
\date{October 2021}

@@

\begin{abstract}
    A large body of literature debates the extent to which media consumption affects consumers' issue attitudes and voting behaviour in contemporary media environments. Studies arguing that the media has minimal influence on opinion formation in today's media environment  have often used designs estimating the effects of novel, previously unconsumed content on participants' attitudes. However, experimental evidence suggests that media consumers discount the bias of different outlets. Additionally, most studies assume rather than measure the extent of newspaper bias. I address both shortcomings with observational evidence exploiting a change in migration framing following an editorial shift in Germany's largest tabloid, \textit{Bild}. The study combines panel data with state-of-the-art methods of text analysis and assesses the effect of changing migration frames in respondents' media diet on their attitudes towards migration. Results across different specifications indicate that even when assessing within-newspaper changes in issue framing, no effect on attitudes or voting behaviour can be found. These findings add further evidence that short-term media effects in the viewers choice era are limited.
\end{abstract}

@@

How does news media content affect citizens' political attitudes and behaviour? Research on political behaviour has discussed this question for nearly a century. Classic theories of agenda setting hold that the attention devoted to issues in the news media will guide citizens to consider these issues important (\cite{McCombs1972}). Similarly, the literature on framing expects that the presentation of issues in the news media will affect recipients' opinions about these issues (\cite{Nelson1997}). Both theories suggest that media coverage strongly affects citizens' understanding of and opinions about politics.

Yet, a critique of these theories maintains that strong news effects are unlikely in the modern media environment, where consumers self-select into news sources of their liking. As a result, "most media users will rarely find themselves in the path of attitude-discrepant information". Even if consumers had to face such discrepant information, this should reinforce existing views rather than persuade as consumers are used to a partisan media diet (\cite[724f]{Bennett2008}). Similarly, agenda setting effects are unlikely as a result of selective exposure (\cite{Lau2021}).

This discussion begs the question: when can we expect media framing effects? Some authors have argued that a lack of media effects can be explained by readers who discount a publications' bias. For example, when a left-wing newspaper endorses a left-wing candidate, this should have no effect on citizens' assessment of the candidate, as the endorsement was to be expected (\cite{Chiang2011a}, see also \cite{Spirig2020}). Additionally, citizens might have developed trust in the position of their preferred outlet and follow changing issue stances of the medium.

@@

@@

To measure the dependent variable (issue attitudes) and differentiate the treatment groups (newspaper readers), I employ panel data from the German Longitudinal Election Study (\cite{GLES2019LongTermTracking}). Following a period of increased refugee arrivals in Europe in the summer of 2015, the 2017 election was dominated by the migration topic and the related upcoming of the radical-right populist challenger party \textit{AfD}. The major two parties dominating German post-war politics, \textit{SPD} and \textit{CDU/CSU}, had to face severe losses following their 'grand coalition' (\cite{Bieber2021, Wessels2021}). This strong (although still variant, see figure \ref{fig:salience}) salience of the migration issue both in the media as well as in citizens subjective evaluations presents an ideal case to study the effects of shifting media framing on issue attitudes, as continuous coverage on and public debate of the issue are a given.

The panel consists of 15 waves, of which 6 (4) contain variables on newspaper consumption, intended voting decision and immigration (integration) attitudes. In each wave 9,000-15,000 respondents were interviewed, which allows the precise estimation of effects of media consumption. Figure \ref{fig:issues} shows the two main dependent variables for each group of readers. Note that higher values indicate more restrictive attitudes towards immigration/integration (see plot subtitles for more information). As can be seen, readers of the tabloid 'Bild' differ strongly from consumers of other outlets, with far more restrictive attitudes. Both variables seem correlated (as expected), with values moving towards the more liberal pole until autumn, where the trend is reversed.

@@

To measure media attention to immigration, I collected 2.5 million newspaper articles from the most important German broadsheets \textit{Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt} and the major German tabloid \textit{Bild} for the period 2013-2019. Table \ref{tab:paper_sales} shows the number of print copies sold and online page impressions for each newspaper per day. The tabloid \textit{Bild} has by far the largest reach, closely followed by the weekly published \textit{Spiegel}. The daily broadsheets sell less, and interestingly the online presence seems to be dominated by the right, given the weak presence of centre-left \textit{SZ} and \textit{taz}.

\begin{table}[!htbp] \centering 
    \begin{tabular}{@{\extracolsep{5pt}} lcc} 
    \\[-1.8ex]\hline 
    \hline \\[-1.8ex] 
    Paper & Print Sales & Web Impressions \\ 
     & (thousands) & (millions) \\
    \hline \\[-1.8ex] 
    Bild & 1756 & 555 \\ 
    Spiegel & 758 & 388 \\ 
    FAZ & 242 & 123 \\ 
    SZ & 358 & 13 \\ 
    taz & 51 & 7 \\ 
    Welt & 521 & 135 \\ 
    \hline \\[-1.8ex] 
    \end{tabular} 
  \caption{Average daily sales and online impressions of different newspapers in 2017. Source: \cite{IVW}. Table generated with stargazer (\cite{stargazer}).} 
  \label{tab:paper_sales} 
\end{table}

To identify whether an article is about migration, I train a BERT deep-learning classifier. As migration content is rare, and my training data needs to be somewhat balanced, I first construct a dictionary of terms related to migration. I use dictionary extension based on German GloVe word-embeddings\footnote{Downloadable from \url{https://deepset.ai/german-word-embeddings}.} to construct a comprehensive dictionary, and apply it to the articles. Based on the relative share of migration words in an article, I draw a stratified sample of 300 articles (100 from the third with fewest migration terms, 100 from the third with most migration terms and 100 from the third in between) for each newspaper for a total of 1,800 papers. Then, a student assistant\footnote{Thank you, Robin. I know it was painful.} hand-coded these articles, assessing whether their main topic is related to migration.

Using this sample, a BERT deep-learning classifier\footnote{\url{https://huggingface.co/bert-base-german-cased}} is fine-tuned on a subset of 1,400 annotated articles. After fine-tuning, the model correctly classifies 95.5\% of the test set (F1: 0.94, recall: 0.93, precision: 0.95). This classifier is then used to annotate all 2.5 million newspaper articles. For the time frame from 2013 to 2019, around 90,000 (3.6\%) are identified to treat the issue of migration.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/salience_papers.png}
    \caption{Share of migration articles out of all articles across time for different newspapers.}
    \label{fig:salience}
\end{figure}

Figure \ref{fig:salience} shows the share of migration articles in each newspaper. Clear peaks in attention are visible surrounding the refugee movements of summer 2015, following the new years eve in Cologne 2016, and the conflict over asylum policy between CDU and CSU in 2018. Newspapers seem to be event-driven and follow largely parallel trends, albeit at different levels, with the  centre-left \textit{SZ} discussing migration most and the tabloid \textit{Bild} least. Attention decreases in mid-2016, but remains high compared to pre-2015 levels. Surrounding the fielding period of the survey waves (gray vertical lines), most newspapers show similar levels of attention to the issue between 2.5\% and 6\% of all articles, with the exception of \textit{SZ}, which moves between 5\% and 12.5\%.

@@

I expect the capital crime, illegal entry/petty crime, and the refugee numbers frames to have negative effects on attitudes towards migrants (as connected to negative considerations regarding immigration), while deportations, internment camps, and drownings in the Mediterranean sea should have positive effects, framing immigration as a humanitarian issue. The labour market frame should show mixed results, given that labour market integration might be perceived as threatening or addressing needs in the German economy. Figure \ref{fig:frames} shows the prevalence of these frames across time for each newspaper. It can be seen that substantial variation exists in the relative emphasis of migration frames both within and across newspapers. Interestingly, some issues seem to vary according to newspaper slant. For example, more right-leaning outlets \textit{Bild, FAZ}, and \textit{Welt} seem to largely emphasise crime and deportations, compared to their centrist and left-wing competitors. This lends some first credibility to the expected effects for these frames, although the differences are not clearly visible for the other frames.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/frames_papers.png}
    \caption{Prevalence of different migration frames across newspapers and time.}
    \label{fig:frames}
\end{figure}

\subsection{Local data on news exposure and voting behaviour}

I am in the process of collecting data on the distribution of different newspapers in German electoral districts, as well as data on voting behaviour within these districts. This allows to assess whether higher readership of \textit{Bild} (and thus more exposure to its migration framing) results in changes in voting behavior following the change in migration framing after the editor-in-chiefs departure (see below) and the subsequent change in migration framing.

\section{Case description and treatment}

During the summer of 2015, the German tabloid Bild responded to the emerging migration issue unexpectedly progressive, calling for support for the newly arriving immigrants and attacking those promoting hate against refugees online (Bild stellt Hetzer an den Pranger 2015; Alvarez 2015), before starting the campaign ’We help - \#refugeeswelcome’ (Grosse Bild-Aktion: Wir helfen - \#refugeeswelcome n.d.). With this move, the paper heavily diverged from its right-wing tradition. The major change in tone was largely ascribed to then editor-in-chief Kai Diekmann, who had also invited a refugee family to live in his house (Reichart 2015). Diekmann left the company in January 2017. Then sole Bild editor-in-chief Julian Reichelt increasingly painted a picture of refugees as criminals and terrorists on the tabloid's front page (Niggemeyer 2018; Zudeick 2018). I aim to exploit this communicative shift in the tabloid’s migration framing to assess the effects of media framing on readers’ issue deﬁnitions, attitudes, and voting behaviour. As figure \ref{fig:frames_diekmann} shows, the migration framing presented in Bild changed following Diekmann's departure. Specifically, immigration was increasingly framed in terms of crime, while other newspapers maintained their levels of attention (see also figure \ref{app:diekmann_all_frames} in the appendix).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/frames_bild_diekmann_pres.png}
    \caption{Monthly attention to different immigration frames following departure of Diekmann from Bild, for Bild and Welt. While attention to other frames follows the Welt's coverage of immigration (or any other newspaper, see appendix \ref{app:diekmann_all_frames}), an increase in Bild attention to general crime news when reporting about immigration is visible following the departure of Diekmann.}
    \label{fig:frames_diekmann}
\end{figure}

Both studies in this paper build on this shift in attention. Study 1 will present estimates from two models assessing whether changing attention to any frame will affect issue attitudes or voting behaviour, using media and opinion data on the migration issue from the German national election in 2017. This is followed by a Difference-in-Difference estimate based on the change in attention to the crime frame presented above. Study 2 will turn to behavioral data and assess whether voting behaviour in voting districts with more circulation of \textit{Bild} have experienced changes in voting behaviour.

@@

I estimate the same DiDs to measure shifts in migration framing in a given newspaper, controlling for shifts in all other newspapers. To do that, I first sum the attention given to a certain frame in a given newspaper in the past day, week, month, or half year, up to the last survey wave. This allows me to assess whether choosing different lags affects my estimates. Then, I estimate the DiD, indicating how much a newspaper changed its attention to a given migration frame, relative to all others. The DiDs estimating the opinion shifts described in \ref{sec:did_readers} are then matched and regressed on the shifts in migration framing. If increasing attention to a certain frame causes a shift in migration attitudes, these two measures should be highly correlated.

Second, I fit two-way fixed-effect models with individual news framing estimates for each survey interview. Since waves are sometimes fielded for several weeks, it might be a strong assumption to treat all interview dates within a given survey wave as identical. I measure exposure to each frame by summing up the frame attention of the past week (relative to the individual interview) in each newspaper the respondent claimed to have read in the past week and dividing it by the total number of newspapers read. This way, I estimate frame exposure for each individual respondent's interview date and news diet, which allows to estimate the correlation of changes in immigration opinion with the migration framing in recently consumed outlets.

While the DiD's are better identified, controlling for shifts in other newspapers and reader groups, they suffer from the fact that surveys are conducted for several days, often up to two weeks, which makes the precise matching of newspaper content impossible. The fixed-effect models control for wave effects and (time-constant) individual characteristics, therefor restricting the explainable variance to everything beyond general trends and individual characteristics, but might include more noise. Another difference is that the DiD-models correlate changes, while the fixed-effect models regress opinion levels on frame attention levels.

\subsubsection{Local changes in voting behaviour}

@@

\subsection{Study 1}

\subsubsection{Correlation of opinion changes among readers with changing frame attention in consumed newspapers}

As figure \ref{fig:issues} shows, and as discussed in appendix \ref{sec:did_ass} and \ref{sec:did_sig}, the dependent variables show highly parallel trends but also display considerable variation beyond statistical noise\footnote{Recent, additional analyses have shown that the opinion shifts vary beyond statistical noise only when considering the broadly defined treatment (all readers of a newspaper). If however consulting only changes among those consuming a single newspaper, significance levels of these shifts approach what would be expected through random sampling error (see figure \ref{fig:did_by_treat}). At this point, it is unclear whether this is an outcome of severely decreased sample size or lack of opinion change among exclusive readers.}
. Can these opinion shifts be explained by shifts in news content? Figure \ref{fig:did_corr} shows the core analysis to estimate the effect of migration framing on reader opinion. The displayed estimates are obtained by regressing the DiD-estimates for issue attitudes on the DiD-estimates for media attention, as described in section \ref{sec:models}. If opinion shifts among newspapers' readers - relative to other readers - can be explained by shifts in newspapers' issue framing, these two variables should be highly correlated.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/vis/effectplot_frames_did.png}
    \caption{Regression coefficients from DiD-models with different specifications.}
    \label{fig:did_corr}
\end{figure}

This is however not the case. Each panel in figure \ref{fig:did_corr} shows the effect of one frame on an attitude variable and the intention to vote for the radical right \textit{AfD} (controlling for all other frames mentioned in section \ref{sec:frame_selection}), across different lags and treatment definitions. Each dot in this graph represents the coefficient of a model explaining standardised\footnote{Meaning that independent and dependent variables were mean-corrected and re-scaled to standard deviations before regression. A coefficient of one means that a one-standard-deviation shift in the independent variable corresponds to a one-standard deviation shift in the dependent variable.} aggregate opinion shifts with standardised shifts in media attention to specific frames.  For example, in the upper left panel 'crime', we can see three different panes ('afd-support', 'immigration' and 'integration') for each dependent variable. The vertical rows distinguish four different lags indicating the frame prevalence of the past day, week, month, half year, or the entire period up to the last survey date. Additionally, the color of the points indicates whether the model was estimated only for readers of a single newspaper (lower N, but no treatment diffusion) or all readers of a newspaper (who might read another newspaper, say SZ readers who also read FAZ).

Only very few estimates are significant and no trends emerge across specifications for any variable. All estimates are clustering around zero, indicating no effect. Estimates of all relationships indicate both positive and negative relationships across specifications. While the 'exclusive reader' estimates show more variation, they also indicate far broader confidence intervals. There is no considerable difference across lags, and the variables only differ by the broader confidence intervals of the integration variable, owing to the fewer waves including this question. Generally, this analysis indicates no effect of media framing of migration on migration attitudes or voting behaviour.

\subsubsection{Fixed-effect model}

Again, most estimates are not significant, differ in direction across specifications, and cluster around zero. While some estimates are unidirectional (crime and labour market on the immigration issue; crime, capital crime, labour market and camps on \textit{AfD}-support), they show tiny effects (a standard deviation increase in attention to the labour market makes immigration attitudes between 0.005 and 0.015 standard deviations more conservative) and are usually not significant. Most importantly, both models do not correspond to expectations. This is probably most visible for the capital and sex crime frame ('capcrime'), which should show strongly positive effects (leading to more restrictive attitudes), but also for the illegal immigration and petty crime frame ('crime'), which consistently shows insignificant negative (i.e. liberalising) effects, contradicting the more sensible increase in \textit{AfD}-support when exposed to this frame.

\subsubsection{Difference-in-Difference estimate}

I initially hoped to match former and current readers of Bild or at least readers of Bild and other outlets to assess whether this shift in framing of a single newspaper affected attitudes and voting behaviour. Regrettably, only one survey wave preceded the departure of Diekmann, precluding a test of the parallel trends assumption. However, even under this strong assumption the estimate might be of interest. At this point, however, I can not present such an estimate yet.

\subsection{Study 2}

Results for local differences in electoral behaviour will be added once the data on local news circulation is available.

This paper assesses the responsiveness of newspaper consumers' migration attitudes to the migration framing of their respective newspaper. Aggregate difference-in-differences models and individual estimates show that consumers' opinions on immigration and integration were not reacting to changes in the newspaper framing of the respective issue. These findings add to an increasing body of evidence showing the absence of short-term media effects in contemporary media environments (\cite{Arceneaux2013, Durante2012, Gentzkow2010, Guess2021, Lau2021}).

Future research will have to establish whether media consumers react to attitude-discrepant information by changing the outlet instead. Several studies found that, instead of reacting with opinion shifts, readers might change their political preferences or mobilise to vote (\cite{Stetka2020}). Similarly, several studies indicate that changes in a media outlet's slant will result in a change in the composition of the consumers (ie those with conflicting attitudes stop reading the newspaper) rather than a change in a fixed readership's attitudes (\cite{Arceneaux2013, Durante2012, Spirig2020}). Whether similar effects can be observed when explicitly measuring newspaper framing are likely next steps for analysis.

If short term changes in framing have no effect on attitudes in the short run, decreases in readership might lead to long-term changes in political opinions due to generational replacement into a changing media environment. This could also make findings from some studies in the field convincingly showing effects (\cite{Foos2020}) consistent with this study and other literature showing weak or absent effects.

Another way forward might be to test theories of \textit{issue evolution} (\cite{Carmines1986}). For example, if a process of 'issue-sorting' took place on the migration issue in 2015, when the topic became very salient, citizens might have chosen a side and stuck to their opinion afterward. Future studies could hence try to identify such windows of opportunity where media effects might still be relevant.

\printbibliography

\subsection{Change in Bild framing of migration issue}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/did_papers_frames_1617_scatter.png}
    \label{app:diekmann_all_frames}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/did_papers_frames_all_scatter.png}
\end{figure}

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{paper/vis/did_papers_allframes_1617_hist.png}
\end{figure}

\subsection{DiD-assumption: Parallel trends}\label{sec:did_ass}

\subsection{Distribution of DiD-effect significance}\label{sec:did_sig}

@@

These highly parallel trends might raise another problem, however. Possibly, no significant variation of reader groups' migration opinion from the general trend can be observed. To address this concern, I estimate DiDs for all wave-issue-paper combinations. This means I use attitudinal questions on all political issues included in the GLES Panel (economic, issues, climate, ...), and assess whether news consumption matters for opinion formation more generally. Figure \ref{fig:p_values} shows the distribution of p-values for the DiD-term from 1,752 fixed-effect models with GLES Panel data. The distribution shows a considerable deviation from the expected uniform distribution (red line) were there no explaining variation in the DiD-term beyond the uniform effects of wave and paper readership. This means, more often than random (12\% vs 5\%), the opinions of consumers of certain news deviate from the opinions of other citizens. This is even more severe for the 192 wave-paper combinations concerning the immigration or integration item used in the remainder of this study, where 18\% of all estimated DiDs fall under the 5\%-threshold\footnote{See plot \ref{fig:p-val_imm}}.  This strongly indicates that news consumption matters for opinion formation and that the data shows variation which can be explained, despite highly parallel trends.

\subsection{Opinion change among all and exclusive newspaper consumers}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/vis/DiD_model_ps_immint_bytreat.png}
    \caption{Empirical p-values of 192 DiD-models showing opinion change for different treatment definitions.}
    \label{fig:did_by_treat}
\end{figure}

@@
